---
title: "Dekut_R"
author: "Dave Karanja"
date: "2024-06-13"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: united
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, include = T)

```

# **CONTROL STRUCTURES AND FUNCTIONS**

# **CONTROL-STRUCTURES**

-   Control structures are used to determine the flow of execution of a program

-   They are used in programming for decision making

-   They are of 2 types:

    1.  Selection statements
    2.  Loops

-   Selection statements include:

    -   $if$ $statements$

    -   $if$ $else$ $statements$

    -   $ifelse()$ function

    -   $Nested$ $if$ $else$ $statements$

    -   $jump$ $statements$

    -   $return$ $statements$

-   There only two types of loops

    -   $for$ $loops$

    -   $while$ $loops$

## 1. Selection Statements

### $if$ $statements$

They test only one condition as either $TRUE$ or $FALSE$

In the event condition tested is $TRUE$ then the statements in the body of the $if$ $statement$ are executed.

Otherwise the program stops

**Syntax**

```         
if(condition){
    statements
    ....
    ....
}
```

[**Click here**](https://media.geeksforgeeks.org/wp-content/uploads/20200331174243/if-statement2.png) to get flowchart

```{r}
# Condition is true thus program is executed
x<-5
if(x>0){
  print("Positive number")
}
```

```{r}
# Condition is false thus program stops
x<- -5
if(x>0){
  print("Negative number")
}
```

### $if$ $else$ $statements$

It tests for one condition as either $TRUE$ or $FALSE$

If condition tested is $TRUE$ then the statements in the body of the $if$ $statement$ are executed, otherwise, the statements in the body of the $else$ $statement$ are executed.

Syntax

```         
if(condition){
    statements
    ....
    ....
}
else{
    statements
    ....
    ....
}
```

[**Click here**](https://www.guru99.com/images/r_programming/032818_1241_IFELSEELIF1.png) for flowchart

```{r}
# Condition is false thus program prints "Negative number"
x<- 1:5
if(x>0){
  print("Positive number")
} else {
  print("Negative number")
}
```

```{r}
x<-runif(1,0,10);x
if(x>3){
  y=10
} else{
  y=0
  
};print(y)
```

```         
runif() creates random variables from the uniform distribution with parameter n, min and max
```

### $ifelse()$ function

The function is used to test for conditions as $TRUE$ or $FALSE$ element wise thus making them effective to work with vectorized statements

Vectorized statements are simply just a vector with at least 2 elements

$if$ $statements$ and $if$ $else$ $statements$ don't test for conditions element wise unless loops are employed to enable the work on vectorized statements.

You can use the $ifelse()$ function to write shorter more readable codes instead. **Syntax**

```         
ifelse(test, yes, no)
test:   Condition

yes :return values for true elements of test.

no  :return values for false elements of test.
```

```{r}
# Using if statement and for loop
a<-c(5,7,2,9)
for(i in seq_along(a)){
  if(a[i]%%2==0){
    print("Even number")
  } else {
    print('Odd number')
  }
}
```

```{r}
# Using ifelse() function
a<-c(5,7,2,9)
ifelse(a%%2==0,"Even number","Odd number")
```

### $Nested$ $if$ $else$ $statements$

These statements are used to test at least 2 conditions

$Nested$ $if-else$ $statements$ allow for more complex decision-making in your code and are useful when you need to evaluate multiple conditions, where the outcome of one condition determines whether to proceed with another condition.

For instance if current condition is $FALSE$ the next condition is tested and so on till $R$ finds a condition which is $TRUE$ and executes it

```{r,eval=F}
num_fact<-as.integer(readline(prompt="Enter a number:"))
fact=1

if(num_fact<0){
  print("Factorial doesn't exist!!")
}else if(num_fact==0){
  print("Factorial of 0 is 1")
  }else{
  for (i in 1:num_fact){
    fact=fact*i
  }
  print(paste("Factorial of",num_fact,"is",fact))
}
```

```{r}
# First condition is true thus its expression is executed
x<- -1
if(x<0){
  print('Negative number') 
} else if (x>0){
  print("Positive number")
}
```

```{r}
# Both conditions are false thus program stops
x<-0
if(x<0){
  print('Negative number')
}else if(x>0){
  print("Positive number")
}
```

```{r}
# First condition is false, R tetsts the next condtion WHich is true thus its expression is executed
x<-1
if(x<0){
  print("Negative number") 
}else if(x>0){
  print('Positive number') 
}
```

## 2. LOOPS

Loops are used to iterate (repeat) given commands without necessarily writing codes for each commands.

In loops we employ the functions $seq$\_$along()$ and $length()$ when we need to iterate through indices of objects.

The functions are very useful when dealing with strings.

Knowledge in indexing is of essence when using the functions in loops.

Unlike $length()$, $seq$\_$along()$ does not consider empty $null$ values as elements and thus does not give them an index position thus making it more effective.

When $seq$\_$along$ finds an empty element it stops the looping process.

However it can't be used in $while$ loops

### $For$ Loops

They perform a finite number of iterations on given commands or operations.

**Syntax**

```         
for(value in vector){
    statements
    ....
    ....
}

value: A variable assigned each element of an object in each iteration
```

[**Click here**](https://media.geeksforgeeks.org/wp-content/uploads/20200327192448/for_flowdiagram.png) *to access flowchart*

Here are some few examples on for loops

```{r,1}
# for loop without seq_along() or length() functions
x <- 6:10
total <- 0
for (i in x) total = total + i
total
```

```{r,2}
# for loop with seq_along() fucntion
x<-6:10
total<-0
for(i in seq_along(x)) total=total +x[i]
total
```

```{r,3}
x<-c("foo","bar","baz")
for(i in seq_along(x)){
  print(i)
  print(x[i])
}
```

```{r}
#fibonacci sequence
fib<-c(1,1)
for(i in 1:10){
  fib<-c(fib,fib[i]+fib[i+1])

}
fib
```

#### Nested $for$ loops

A nested loop is a loop that is within another loop

```{r,4}
for(i in 1:3){
  for(j in 1:2){
    print(c(i,j))
  }
}
```

**Explanation of chunk above:**

An outer loop is created that iterates through the values 1,2,3 and assigns them to the variable $i$

For each value assigned to $i$ an inner loop is created that iterates through the values 1,2 and assigns them to a variable $j$\
The concatenation function $c()$ combines i and j as vectors

```{r}
m<-matrix(2:15,nrow=2);m
for(r in 1:nrow(m) ){
  for(c in 1:ncol(m)){
    print(m[r,c])
  }
}
```

### $while$ loops

They are used to iterate through commands until a certain condition is met.

It's used when one is unsure of the number of iterations they want to perform.

The loop runs until the condition tested is FALSE

While loops must have an iteration function

**Syntax**

```         
while(condition){
    statement
    ....
    ....
}
```

[**Click here**](https://media.geeksforgeeks.org/wp-content/uploads/20200327192542/while_flowdiagram.png) for flowchart

Disclaimer: If one does not define their conditions correctly, it may lead to infinite loops.

```{r}
x<-2:20
s=0
i=1
while(i<=length(x)){
  y=x[i]
  i=i+1
  s=s+y
  print(s)
}
```

```{r}
x<-letters[4:10]
i=1
while(i<=length(x)){
  y=x[i]
  i=i+1
  print(y)
}
```

```{r}
fib<-c(1,1)
i=0
while(i<=10){
  fib=c(fib,fib[i]+fib[i+1])
  i=i+1
}
fib
```

### $jump$ $statements$

Jump statements are used to terminate loops at a particular iteration or skip a particular iteration in the loop

These statements include:

-   Break statement

-   Next statement

Jump statements require a defined condition to tell them when to terminate or skip a particular iteration

#### $break$ $Statements$

It is used to terminate a loop at a particular iteration.

After termination the program continues with the next statements outside the loop **Syntax**

```         
if (condition) {
break
}
```

[**Click here**](https://media.geeksforgeeks.org/wp-content/uploads/20191120193634/Break-statement-python.jpg) for flowchart

```{r}

for (i in c(3, 6, 23, 19, 0, 21))
{
	if (i == 0)
	{
		break
	}
print(i)
}
print("Outside Loop")

```

#### $next$ $statement$

It discontinues a particular iteration and jump to the next iteration. **Syntax**

```         
if (ondition) 
{
    next
}
```

```{r}

for (i in c(3, 6, 23, 19, 0, 21))
{
	if (i == 0)
	{
		next
	}
	print(i)
}
print('Outside Loop')

```

### $repeat$ $statement$

Just like the $while$ $loop$ the $repeat$ $statement$ is used to iterate through commands when there's uncertainty of the number of times repetition should be done.

But unlike the while loop, it does not have a condition to be satisfied in order to terminate the loop.

Instead it employs the $break$ $statement$ termination of the loop is required

**Syntax**

```         
repeat { 
   statements
   ....
   .... 
   if(condition) {
      break
   }
}
```

[**Click here**](https://media.geeksforgeeks.org/wp-content/uploads/20200327192638/repeat_flowdiagram.png) for flowchart

```{r}
x=0
cat("Odd numbers:","\n")
repeat{
  x=x+1
  if(x%%2==0){next}
  if(x>10) {break}
  cat(x);cat(" ")
  };cat("\n")
```

# **Functions**

A function is a predetermined formula that takes inputs and gives output. $f(x)$=x^2^

A formula is a mathematical expression that shows the relationship between variables. $A$= πr^2^

$R$ is a functional language.

Thus it's better to use recursive functions rather than loops.

Functions are of two types:

-   In built functions

    -   They are already defined in $R$

-   User defined Functions

    -   They are defined and declared by the user.

## In Built Functions

### The $apply$ family

The apply functions are used to apply functions to objects and they include:

-   $apply()$

-   $lapply()$

-   $sapply()$

-   $vapply$

#### 1. $apply()$

Maps functions to objects that are 2 dimensional

Best explained by a matrix

**syntax**

```         
apply(X, MARGIN, FUN)
X:object
margin : 1 rep row, 2 rep column
FUN: Function to be mapped
```

```{r}
m <- matrix(1:9, nrow = 3, byrow = TRUE);m
print(m)
#Apply sum function to each row
(row_sums <- apply(m, 1, sum))
print(row_sums)
# Apply sum function to each column
col_sums <- apply(m, 2, sum)
print(col_sums)
```

#### 2. $lapply()$

Maps functions to list.

-   It takes a list as input and returns a list as an output

**Syntax**

```         
lapply(X, FUN, ...)
```

```{r}
(l<-list(1,2,3))
lapply(l,function(x) x^2)
```

**NB: If elements of list are named, the function mapped is not applied to this names**

```{r}
(l<-list(a=1,b=2,c=3))
lapply(l,function(x)x^2)
```

-   If given any other object ie: a vector as its data, it will still output a list

```{r}
lapply(1:3,function(x)x^2)
```

-   If elements of the list are more complex than a single number, the function is still applied to the element of each list

```{r}
k<-list(a=1:3,b=4:6)
lapply(k,function(x)x^2)
```

#### 3. $sapply$

It maps functions to a given object and ttys to determine the best output to return, either, a vector, list, matrix, e.t.c

Often it returns a vector but if it fails to determine output it returns a list

```{r}
sapply(1:3, function(x)x^2)

sapply(k, function(x)x**2)
```

```{r}
sample_data<- data.frame( x=c(1,2,3,4,5,6), 
                          y=c(3,2,4,2,34,5)) 
sapply(sample_data, max) 
```

```{r}
m_v <- c(1, 2, 3, 4, 5)
sapply(m_v, function(x) if (x %% 2 == 0) c(x, x^2) else c(x))
```

```         
Because output of chunk bove have different lengths, sapply() returns the outputs as a list
```

#### 4. $vapply()$

## User Defined Functions

**Syntax**

```         
function_name = function(arg_1, arg_2, …) 
{ 
Function body 
} 

function_name(arg_1, arg_2, …)

function_name: Name of function that is saved as an object in the R enviroment

arg_1,arg_2, ... : They are arguments of the function which are simply placeholders(Rep input function takes when called)

Function body: Contains statment that define what function does

function_name(arg_1, arg_2, …):Refers to calling the function for execution.
```

#### Various ways of calling functions

1.  **Using position and name of argument**

Here, the order of arrangement of function matters as it determines which argument gets to be assigned which value

```{r}
div<-function(x,y){
  print(paste(x,"/",y,"is",x/y))
  x/y
}
div(6,2)
div(2,6)
```

To avoid following the order of arguments when calling a function we can define the values of the arguments when calling it.

This is only possible for named arguments

```{r}
div(y=2,x=6)
```

2.  **Calling functions without arguments**

In order to do this, no arguments are defined when declaring a function

```{r}
no_arg<-function(){
  for(i in 1:5){
    return(i^3)
  }
}
no_arg()
```

3.  **Calling functions with Default Arguments**

Here, the values of arguments are defined when declaring the function

This enables calling of the argument without arguments

It also enables one call an argument with arguments by defining its values during the call. This overwrites values of argument defined during declaration

```{r}
roots<-function(a=1,b=-3,c=2){
  r_1 = (-b-sqrt(b^2-4*a*c))/2
  r_2 = (-b+sqrt(b^2-4*a*c))/2
  if(r_1&r_2<0){
    r = -b/(2*a)
    return(c(r,r))
  }else if(r_1&r_2==0){
    return(0)
  }else{
    
    return(c(r_1,r_2))
  }
}
roots()
roots(a=1,b=-2,c=1)
```

#### Lazy Evaluation of Functions

This simply implies that arguments of a function are evaluated when need arises to

```{r}
f<-function(x,y=x^2) {
  1+x
}
f(2)
```

```         
In the hunk above, arg y is not evaluated as its not needed. It will only be evaluated when need arises. Check next Chunk
```

```{r}
f<-function(x,y=x^2) {
  y+x
}
f(2)
```

Lazy evaluation allows for creation of functions that have default arguments which are dependent on other arguments defined in the function

### Types of Functions

#### Recursive Functions

These are functions that call themselves.

When the function is called it calls itself again and again and in the process it breaks down a given problem into small finite sizes

This leads to formation of a loop

Recursive functions break down a problem into small bits then merge the output of the small bits to give output of the parent problem

```{r}
a<-function(k){
  if(k>0){
    result=k+a(k-1)
    print(result)
  } else{
    result=0
    return(result)
  }
}
a(6)
```

The chunk above works as a loop and the table below shows its break down

| a(k) | Condition | result  | print(result) |
|:----:|:---------:|:-------:|:-------------:|
| a(6) |   TRUE    | 6+a(5)= |    6+15=21    |
| a(5) |   TRUE    | 5+a(4)= |    5+10=15    |
| a(4) |   TRUE    | 4+a(3)= |    4+6=10     |
| a(3) |   TRUE    | 3+a(2)= |     3+3=6     |
| a(2) |   TRUE    | 2+a(1)= |     2+1=3     |
| a(1) |   TRUE    | 1+a(0)= |     1+0=1     |
| a(0) |   FALSE   |   0=    |       0       |

1^st^ 3 cols are filled row-wise and the last col filled column wise

Here's a Fibonacci sequence using recursive functions

```{r}
fib_recursive <- function(n) {
  if(n == 1) {
    return(1)
  } else if(n == 2) {
    return(c(1, 1))
  } else {
    previous_fib <- fib_recursive(n - 1)
    return(c(previous_fib, sum(tail(previous_fib, 2))))
  }
}
fib_recursive(10)
```

#### Scoping

#### Closures

$This$ $is$ $the$ $furthest$ $we$ $can$ $get$ $with$ $functions$ $for$ $today!!!$

$As$ $I$ $wish$ $you$ $all$ $the$ $best$ $in$ $your$ $journey$ $in$ $R,$ $I$ $want$ $to$ $thank$ $you$ $for$ $your$ $attendance,$ $patience$ $and$ $attention$ $throughout$ $the$ $session.$
