---
title: "DekutR"
author: "Gathu Macharia"
date: "2024-04-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: united
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include= F}
knitr::opts_chunk$set(echo = T,message = F,warning = F, include=T)
```

# Creation of Project Folders.

Reproduce your work with renv\
Keeps R History (Global environment is stored)\
Loading data is an easy task (No need for path! if you do it right)

An R project enables your work to be bundled in a portable, self-contained folder. Within the project, all the relevant scripts, data files, figures/outputs, and history are stored in sub-folders and importantly - the working directory is the project's root folder. You cant appreciate this enough until you start working on big projects.

# Reproducible Programming

Use R markdown or sweave. We use rmarkdown for our case.

Repoducability\
Speed of task management\
Smart workflow

## Reproducible Environment

$renv::init()$\
$renv::activate()$\
$renv::snapshot()$\
$renv.restore()$\
to check the version of package.\
package Version("ggplot2")

## Git

future lesson!

# Expressions

## Arithmetic Expressions

`+`\
`-`\
`*`\
`^` exponent `sum` sum `/` is used in division

`%%` is used in modulus division- Returns the remainder

`%/%` used in integer division- Returns integers from divisions getting rid of decimal/floats

```{r}
 1+2

4%%3

4%%2
4%/%2

4%%3

5%%3
```

## Boolean Expressions

## Comparison Operators

`==` Equals to

`!=` Not Equal to

`\>`

`\<=` less than or equal to They are expressions whose output is either true or false.

## Logical Operators

`|` OR `&` and `!` Logical operator not that negates values

`|` logical element wise OR operator- Returns TRUE if at least one condition is TRUE

`&` logical element wise AND operator - Only returns true if all are TRUE

```{r}

4>6
4<=6


4<6 & 4>3
4<6 & 4<3
!TRUE
!FALSE
x<-c(T,F,T,F)
y<-c(T,T,F,F)
x|y

x&y
```

# Variable Assignments

```{r}
m<-c('mango',"Apple","Berries")
length(m)

m=c('mango',"Apple","Berries","Mary's")
m

c(1,56,56)->z

z[1]<-2
z 

#Other specific assignments for R environments!
```

```{r,echo=F,include=F}
###### another way of assignments #############
c(1,2,7,9,10,33)->z
z[1]
z[1]<-2
z

```

# Basic Data Types

-   The `class()` is used to check the data type of data

## 1. Numeric Data

-   It is data whose values are members of real number

-   Such as :

    -   Integers(whole numbers)

    -   Double/Floats/Decimals

    -   Fractions

    -   Complex/imaginary numbers

-   We can coerce data to Numeric using the `as.numeric()`

-   We can check if numeric using the `is.numeric()`

```{r}
is.numeric(2)
class(2)
```

## 2. Integer data

-   It is data whose values are whole numbers

-   By default R saves all numbers as Numeric.

-   To coerce a number as Integer, we add a prefix `L` to it

-   We can also use the `as.integer()`

-   When a decimal is coerced as integer it loses its floating numbers

```{r}
is.integer(2)
is.integer(2L)
x<-as.integer(2)
class(2L)
class(x)
as.integer(2.8)
```

## 3. Complex Data

-   Refers to data that is imaginary

-   Imaginary implies that the data is not a member of the real numbers

-   A number is defined as complex if it has a prefix `i`

```{r}
1+0i
is.complex(1+0i)
class(1+0i)
```

-   Dealing with complex data enables us find the square root of negative numbers

```{r}
sqrt(as.complex(-5))
```

## 5. Character Data

-   Refers to data whose values are strings

-   Any data value inside double `""` or single `''` quotations marks is regarded as a string

```{r}
x<-"hello world"
class(x)
is.character(x)
as.character(3.14)
```

# Data Structures

-   By concatenating simple data types, you create data structures and they include

## 1. Vectors

-   A vector is a one dimension data structure that stores data of the same type

-   Vectors are also called atomic sequences

```{r}
v<-c(1,2,3)
1:3
rep("foo",3)
```

-   We can check if a data structure is a vector using the `is.atomic()` or `is.vector()` functions

```{r}
is.atomic(v)
is.vector(v)
```

-   In the event a vector has an attribute then `is.vector()` will return FALSE

-   The `is.atomic()` returns TRUE confirming that its a vector even though it has attributes

```{r}
v<-1:3
is.vector(v)
attr(v,"foo")<-"bar";v
is.vector(v)
is.atomic(v)
```

-   In the event you have different data types in a vector, R tries to coerce them to the most suitable data type and it follows the below priority

    -   Character

    -   Complex

    -   Numeric

    -   Integer

    -   Logical If for example we have a vector of all data types.

    ```{r}
    x<-c(1,2,3i,"foo",2L,2<3)
    class(x)
    x<-c(1,2,3i,2L,2<3)
    class(x)
    x<-c(1,2,2L,2<3)
    class(x)
    x<-c(2L,2<3)
    class(x)
    x<-2<3
    class(x)
    ```

## 2. Matrix

-   It is a 2 dimensional data structure that stores data of the same type

-   A matrix is a 2 dimensional vector

    -   The `dim()` is used to define the dimensions(rows and columns) of a vector to make it a matrix

    -   The `attribute()` function is used to check the dimensions of a matrix

```{r}
v<-1:6
attributes(v) #returns null as v has no dimensions yet
dim(v)<-c(2,3) #We've defined the dimensions of v making it a matrix with 2 rows and 3 columns
attributes(v) # checks dimensions
dim(v) # checks dimensions
v # is now a matrix with 2 rows and 3 columns
```

-   By default values in a vector r added column-wise by R

-   The `matrix()` is used to create matrices and can be used to define if data is to be input row wise or column wise

-   The argument `byrow=T` indicates that data is input row-wise into a matrix

```{r}
v<-1:6
v<-matrix(v,nrow=2,ncol=3,byrow=F);v
v<-1:6
v<-matrix(v,nrow=2,ncol=3,byrow=T);v
```

-   **NB:** In Matrices , `*` is used for scalar multiplication while `%*%` is used for matrix multiplication

```{r}
(a<-matrix(1:4,nrow=2))
(b<-matrix(5:8,nrow=2))
a*b
a%*%b
```

-   `t()` is used to transpose a matrix while the `solve()` is used to inverse a matrix

```{r}
t(a)
solve(a)
solve(a)%*%a
```

## 3. Lists

-   Lists and Arrays store data in layers

-   A list is a one dimension data structures that stores different data types in layers

-   An Array is a 2 dimensional data structure that stores different data types in layers

-   A Data frame is a 2 dimensional data structure that stores different data types

-   The `list()` function is used to create lists

-   

```{r}
list(1:3,5:8) # A list of 2 vectors 
list(1:3,c(T,F))
```

-   We can also create **recursive lists** (A list inside another list )

```{r}
list(list(),list(list(),list()))
```

```{r}
(list_s<-list(c("jan","feb",'march','april'),
             matrix(c(3,9,5,1,-2,8),nrow = 2),
             list("flowers",'chocolate')))
```

-   The `unlist()` is used to reduce a list into a vector

```{r}
unlist(list(1:4,5:7))
```

### Indexing

The `?"[["` displays documentation for indexing in R

```{r}
?"[["
```

-   We use indexing to access elements of different data structures

1.  **Indexing Vectors**

```{r}
v<-1:4
v[2]
v[2:3]
# Special case for using vector of indices
v[c(1,1,4,3,2)]
v[-1]
v[-(1:2)]
```

-   You can't combine negative and positive indices

-   Another way to index is by using Boolean expressions or a Boolean vector

-   The Boolean vector should be of the same size as the vector

```{r}
v[v%%2==0]

```

-   By Indexing you can assign values to a vector or add new values

```{r}
v[v%%2==0]<-13;v
```

2.  **Indexing matrices**

```{r}
m<-matrix(1:6,nrow=2,byrow=T);m
m[1,] #access 1st row
m[,1] # access 1st column
```

-   When accessing a single row or column, it is reduced to a vector

-   Using the argument `drop=F` we maintain the single rows and columns as 2 dimensional

```{r}
m<-matrix(1:6,nrow=2, byrow=T);m
m[1,,drop=T]
m[,1,drop=F]
```

3.  **Indexing Lists**

-   **NB:** When sub-setting list using indexing`[]` we get another list

-   This case applies even if you're sub-setting a single element

    -   Reason: When sub-setting a single element your'e not getting that elemnt but the list that contains it

```{r}
L<-list(1:3,6:7)
is.list(L[1])
L[[1]]
```

-   To get that element instead, you have to use double square brackets

```{r}
L[[1]]
```

### Named values

-   Elements in vectors or lists can have names

-   These are the attributes that don't affect the values of the elements and can be used to refer to them

```{r}
v<-c(a=1,b=2,c=3,d=4);v
l<-list(a=1:5,b=c(T,F));l
```

-   The `names()`function can also be used in naming the elements

```{r}
names(v)<-LETTERS[1:4];v
```

-   The names of this elements can be used to Access them

-   This is referred to as string indexing

```{r}
v["A"]
l['a']
l[["a"]]
```

-   When elements are named, one can use the `$` to access the elements

-   Essentially it works like the `[[]]` But there's no need of having the names in quotations

-   **NB: `$`** can't be applied to vectors

```{r}
l$a
```

-   When `[[]]` is used in vectors, it allows you to only extract one element and if the element is name, the name is removed

```{r}
v
v[[1]]

```

## 4. Factors

Factors in R are a way to handle categorical data. Categorical data is data that can be divided into a limited number of categories or groups. These categories are known as levels.

Imagine you have a survey where people can rate their satisfaction as "low", "medium", or "high". These ratings are categorical data because there are only a few possible values (levels) the rating can take.

## Nominal Categorical Variables

Order does not matter $as.factor(x)$

```{r}
color_vector <- c('blue', 'red', 'green', 'white', 'black', 'yellow')
color_vector_f<-as.factor(color_vector)
class(color_vector_f)
is.factor(color_vector_f)
```

## ordinal

Order matters Ordinal categorical variables do have a natural ordering. We can specify the order, from the lowest to the highest with order = TRUE and highest levs="desired order".

```{r}
day_vector <- c('evening', 'morning', 'afternoon', 'midday', 'midnight', 'evening')
# Convert `day_vector` to a factor with ordered level
factor_day <- factor(day_vector, order = TRUE, levels =c('morning', 'midday', 'afternoon', 'evening', 'midnight'))
class(factor_day)
str(factor_day)
##descending to order
factor_day <- factor(day_vector, order = T, levels =rev(c('morning', 'midday', 'afternoon', 'evening', 'midnight')))
```

# Dataframe

Tabular Structure: Organizes data into rows and columns. Variable Heterogeneity: Accommodates different data types in one data set. Integration: Works seamlessly with many R packages and functions. Data Transformation: Facilitates easy manipulation and transformation of data. Compatibility: Imports from and exports to various data sources. Interactive Analysis: Supports interactive exploration and visualization. Reporting: Enables generation of formatted reports and summaries.

```{r}
# Create a data frame
Data_Frame <- data.frame (
  Training = c("Strength", "Stamina", "Other"),
  Pulse = c(100, 150, 120),
  Duration = c(60, 30, 45)
)

# Print the data frame
Data_Frame
#Adding New columns
Data_Frame$school<-c('a','b','c')

```

## Acessing data frames

```{r}
Data_Frame[1]

Data_Frame[["Training"]]


#### We often use this####
Data_Frame$Training
```

## Manipulate

```{r}
# add a column
Data_Frame$heartrate=Data_Frame$Pulse/Data_Frame$Duration
Data_Frame
# append a column
Data_Frame2=data.frame(volume=sample(3000:5000,size = 3,replace = T))
new_dataframe=cbind(Data_Frame,Data_Frame2)
new_dataframe

#### append rows
new_dataframe2=rbind(Data_Frame,Data_Frame)
new_dataframe2
```

```{r}
Data_Frame_New <- Data_Frame[-c(1,3), -c(1,2)]

# Print the new data frame
Data_Frame_New
ncol(Data_Frame_New)

```

$ncols()$

$nrows()$

$dim()$

$colnames()$

# Question

Write matrix $A=5\times 5$ and $B=5\times 5$ matrix. Carry out both scalar multiplication and matrix multiplication, assign the two values $scalar\_a$ and second matrix $mat\_a$. Find out if the object in the third row and second column in$mat\_a$ is less than object in fourth row and second column $scalar\_a$.

```{r,echo=F,include=FALSE}
a=matrix(1:16,ncol=5,nrow = 5)
b=matrix(seq(1,50,by=2),ncol = 5,nrow = 5)
scalar_a=a*b
mat_a=a%*%b
mat_a[3,2]<scalar_a[4,2]
```

\vspace{2.5cm}

If you have come all this way [**click here**](https://www.youtube.com/watch?v=QWEcMqM0acU&ab_channel=Memebyme)

For those who got the answers [**click here**](https://www.youtube.com/watch?v=lPk_zyRKs1Q&ab_channel=Lilmaina)
