---
title: "R Notebook"
output: html_notebook
---

# stringr and forcats packages
# stringr package
## Introduction
* The stringr package provides you with a collection of functions for working with strings.
* It's consistent with function naming, which isn't always given in other packages. For example, all stringr functions have a prefix of str_, followed by the function name.
* You can expect to find pretty much any function you can imagine, from simple string operations to pattern matching, substitution, trimming, splitting, and much more. 

## Installing the package
```{r}
install.packages("tidyverse")
```

Alternatively
```{r}
#install.packages("stringr")
```

## stringr in Action 
## 10 Functions to Preprocess Textual Data
### Declare a vector x that contains five strings

```{r}
library(stringr)

x <- c("house", "car", "plant", "telephone", "arm chair")
print(x)
```

### 1. str_length()
The `str_length()` function returns the number of characters in a given string. When applied to a vector, it returns a vector with each item representing the number of characters in the corresponding string. It accepts a string or a vector as a parameter and returns an integer or a vector of integers, respectively.

```{r}
str_length(x)
```

### 2. str_sub()
The `str_sub()` function returns a substring of a given string. It takes three parameters:

1. The string (or a vector of strings)
2. The starting index of the substring
3. The ending index of the substring

It's useful when you want to limit the number of characters or trim the start/end of a string.

#### Indexing:
```{r}
# Define the vector
words <- c("house", "car", "plant", "telephone", "arm chair")

# Split each word into its individual characters and create a list of characters
split_words <- strsplit(words, "")

# Function to get the index of each letter
index_letters <- function(chars) {
  return(paste(chars, 1:length(chars), sep = ":", collapse = " "))
}

# Apply the function to each item in the list
indexed_words <- sapply(split_words, index_letters)

# Print the result
print(indexed_words)

```
#### Example
```{r}
str_sub(x, start = 2, end = 5)
```

### 3. str_detect()
This function returns a boolean or a vector of booleans. The value depends on whether the entered pattern exists in a given string or not.

The `str_detect()` function takes two parameters - your string (or vector of strings) and a pattern to search for. If the pattern is found, the function returns TRUE; otherwise, it returns FALSE.

```{r}
str_detect(x, "ar")
```
select only those input strings that satisfy the condition:
```{r}
x[str_detect(x, "ar")]
```

### 4. str_replace()
The str_replace() function is useful when you want to replace the first occurrence of a pattern in a string with a specified replacement string. It takes three parameters:

1. The string (or a vector of strings) to search
2. The pattern to search
3. The replacement string

The function returns a modified string in which the pattern to search is replaced with the replacement string, but only at the first occurrence.

```{r}
str_replace(x, "e", "***")
```

### 5. str_replace_all()
It replaces all occurrences of the search pattern with the provided replacement string.
```{r}
str_replace_all(x, "e", "***")
```
### 6. str_count()
The `str_count()` function is here to count the number of times a search pattern appears in a string. It takes two parameters: the string on which the search is performed (or a vector of strings), and the search pattern which can also be a regular expression.

This function will return an integer (or a vector of integers) representing the number of times the search pattern was found.
```{r}
str_count(x, "a")
```

### 7. str_subset()
The str_subset() function returns a subset of a vector of strings that match a certain search pattern. It takes in two parameters: the vector of strings to search and the search pattern itself.
```{r}
str_subset(x, "a")

```
### 8. str_trim()
The `str_trim()` function is useful when you have messy strings full of leading and trailing whitespaces. It will remove all of them, either from a single string or from a vector of strings.

```{r}
y <- c("  hello ", "from  ", " R   ")
print(y)
```
### pass this vector into the str_trim() function
```{r}
str_trim(y)
```
#### 9. str_split()
This function will split a string or a vector of strings into a vector of substrings or a list of vectors of substrings, depending on the format of data passed in.


```{r}
z <- c("office chair", "front desk", "brown laptop case")
print(z)
```
We can now call str_split() on z and pass in space as a delimiter:
```{r}
str_split(z, " ")
```

### 10. 10. str_to_xyz()
There’s actually no function named str_to_xyz(), but there’s a set of functions for transforming a string or a vector of strings. You can use one of the following functions:

* `str_to_title()` – To capitalize first letter of each word in a string
```{r}
str_to_title(x)
```

* `str_to_sentence()` – To capitalize the first letter of a string

* `str_to_upper()` – To uppercase the entire string
```{r}
str_to_upper(x)
```


* `str_to_lower()` – To lowercase the entire string



### 11. str_match(x, pattern) 
extracts parts of the match defined by parentheses:
The code extracts 3 characters where the middle one is a vowel
```{r}
# extract the characters on either side of the vowel
str_match(x, "(.)[aeiou](.)")
```


***REFERENCE***

* https://stringr.tidyverse.org/

* https://stringr.tidyverse.org/articles/from-base.html

* https://medium.com/appsilon-data-science/stringr-10-examples-on-how-to-do-efficient-string-processing-in-r-78d859e25300

# forcats package
R uses factors to handle categorical variables, variables that have a fixed and known set of possible values. Factors are also helpful for reordering character vectors to improve display.

`fct_reorder()`: Reordering a factor by another variable.
`fct_infreq()`: Reordering a factor by the frequency of values.
`fct_relevel()`: Changing the order of a factor by hand.
`fct_lump()`: Collapsing the least/most frequent values of a factor into “other”.

## Installation:
### install the whole tidyverse:
```{r}
#install.packages("tidyverse")
```


### Alternatively, install just forcats:
```{r}
#install.packages("forcats")
```


### Or the the development version from GitHub:
### install.packages("pak")
```{r}
#pak::pak("tidyverse/forcats")
```

## Getting started
### Load the necessary libraries for data manipulation and visualization
```{r}
library(forcats)
library(dplyr)
library(ggplot2)
```

### Count the number of characters for each species, excluding those with missing species
```{r}
starwars %>% 
  filter(!is.na(species)) %>%
  count(species, sort = TRUE)
```

### Group less common species into an "Other" category and count the occurrences of each species
```{r}
starwars %>%
  filter(!is.na(species)) %>%
  mutate(species = fct_lump(species, n = 3)) %>%
  count(species)
```

###  Create a bar plot of eye color frequencies, flipping the coordinates for better readability

```{r}
ggplot(starwars, aes(x = eye_color)) + 
  geom_bar() + 
  coord_flip()
```

### Reorder eye colors by their frequency and create a bar plot, flipping the coordinates for better readability
```{r}
starwars %>%
  mutate(eye_color = fct_infreq(eye_color)) %>%
  ggplot(aes(x = eye_color)) + 
  geom_bar() + 
  coord_flip()
```

### Reorder eye_color based on the number of occurrences and plot
`fct_reorder()` is useful when you want to reorder factor levels based on the values of another variable. However, for the bar plot of eye colors, we can simulate its usage by reordering the eye colors based on their frequency.
```{r}
starwars %>%
  filter(!is.na(eye_color)) %>%
  mutate(eye_color = fct_reorder(eye_color, -table(eye_color)[eye_color])) %>%
  ggplot(aes(x = eye_color)) + 
  geom_bar() + 
  coord_flip() +
  labs(title = "Bar Plot of Eye Colors (Reordered by Frequency)",
       x = "Eye Color",
       y = "Count")
```
### Manually reorder eye_color and plot
`fct_relevel()` allows you to manually specify the order of factor levels. Let's say you want to prioritize some specific eye colors at the top of your plot.
```{r}
starwars %>%
  filter(!is.na(eye_color)) %>%
  mutate(eye_color = fct_relevel(eye_color, "blue", "brown", "green", "yellow")) %>%
  ggplot(aes(x = eye_color)) + 
  geom_bar() + 
  coord_flip() +
  labs(title = "Bar Plot of Eye Colors (Manual Reordering)",
       x = "Eye Color",
       y = "Count")

```





***REFERENCE***

* https://medium.com/@kierisi/r4ds-week-8-factors-forcats-4b7f27cffd60

* https://forcats.tidyverse.org/

